Efficient pedigree recording for fast population genetics simulation

### 2018-01-11 forwards_paper.tex
First pass moving simupop

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

To demonstrate the method's flexibility and to test it in different contexts,
we implemented these simulations separately using two different simulation engines:

\jda{Note/TODO: the varying references to c++, python, fwdpp, fwdpy11 and
	simupop are probably confusing to a new reader. This is in part because
	we refer to the python/simupop method in various places as one or the
	other; the same goes for c++/fwdpy11/fwdpp. We should decide what to
	call these and make our usage consistent}

\begin{enumerate}

    \item
        In \cpp{}, using \fwdpp{} library functions and interface code (details below),
        using a continuum-sites model for both mutation and recombination. Simulations were run using \fwdpy{}, a Python
        package based on \fwdpp.
        % $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
        % We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.

    \item
        In Python, using the \simupop{} library interfacing with Python code
	(via our package \ftprime{}),
        using a discrete but large number of sites.
        % $\mu$ and $r$ define per-base-pair mutation rate and we
        % defined the length of the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.
        % We varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

\end{enumerate}

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  With \fwdpy{}, simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
With \simupop{}, we were only able to run simulations with $N=10^3$ and region sizes up to $10^4$.  Pedigree tracking
dramatically improves run times for both \fwdpy{} and \simupop{} (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
(Figure~\ref{fig:relative_speedup_selection}).  For \fwdpy{}, pedigree tracking reduces runtimes more for larger
population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).  We saw a similar trend with $4Nr$ for \simupop{}, but the effect of $N$
could not be explored because the simulations tended to run out of available memory.
Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when tracking
pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative
behavior with a numerically larger benefit to pedigree tracking (see~\ref{ss:timing_nosel}).

In our test runs, our method used substantially more RAM than simple forward simulations (see~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see~\ref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
        different diploid population sizes ($N$). The left column shows run times for
        simulations including the tracking of neutral mutations.  Missing data points are due to either simulations
    being killed at 72 hours or running out of memory (which only happened with \simupop{}).  The right column of
panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
the results for an implementation using \fwdpy{} where the pedigree simplification steps were handled in a separate thread of
execution and fitness calculations were parallelized.}
\end{figure}

@@

% Comparison of simulation with/without msprime, using \simupop{}
% or maybe just a simple haploid simulation with 1000 QTL and stabilizing selection on a trait (say).
%
% Maybe an estimate of how long \emph{just} the pedigree recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$ whole chromosomes for $10^7$ generations
% in a day.

@@

%%%%%%
\subsection*{Implementation overview}

@@

The \msprime\ Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences. Interoperation with Python simulators
is then straightforward. For the proof-of-concept implementations discussed
here, we took two different approaches to doing this. In the first method,
based on \simupop{}, the simulation engine invokes callbacks at critical points
of the simulation, and we infer nodes and edges from the information that is
provided. Rows are appended to the tables one-by-one, and the tables are
periodically sorted and simplified to control memory usage. The second method we developed uses
\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.

While these methods work well, and indeed result in significant performance
gains over non-ancestry based simulation, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in C or C++ out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime\ C library directly as part of
the core simulation logic, in the manner outlined in Algorithm W. If designed
in this way, the simulation engine would not need to keep copies of the node
and edge information and would not need to infer edges;
even greater gains in computational efficiency could then be obtained. To
facilitate this usage, a standalone, embeddable C library encompassing the core
tree sequence functionality is planned, which should make recording and
outputting tree sequences from any C/C++ code straightforward.

@@

\subsection*{The \cpp{} implementation}

@@

\subsection*{The \simupop{} implementation}

We implemented simulations in a Python script, usable as a command-line utility,
taking input parameters for population size $N$ and scaled recombination rate $\rho$.
As described in 'Simulation benchmarks', we set scaled mutation rate $\theta=\rho$,
we used simplify interval of 1000 and simulated for $T=10N$ generations in each
simulation, introducing deleterious mutations at a rate of $\rho/100$ per
generation.
Because \simupop{} requires explicit locations for loci, we set the number of
loci $n=\rho / (4Nr)$ and marked
$k={\rm round}(\mu / (100 (\mu + \mu / 100))) = 0$ of these as selected loci.
\jda{unfortunately probably what we did...}
The selected loci were chosen as a random sample of all the loci.

@@

The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel} and for the
simulations without selection in Figure~\ref{sfig:ramnosel}.  For our parameters, which involved simplifying every
$10^3$ generations,  pedigree tracking requires more RAM with \fwdpp{} and less
RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is clear.  When pedigree tracking, a non-recombinant
individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.

@@

\begin{figure}
    \includegraphics[]{sims/memuse_nosel}
    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
\end{figure}

@@

We performed a limited set of simulations with \fwdpy{} to explore the effect of the simplification interval on run times and memory
use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
often.

@@

We also performed a more limited set of simulations without natural selection.  The total run times are shown in
Figure~\ref{sfig:rawspeed_nosel} and show the same qualitative behavior as simulations with selection
(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
and~\ref{sfig:speedup_nosel}.

@@

\begin{figure}
    \includegraphics[]{sims/speedup_nosel}
    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
neutral variation over the run times tracking the pedigree.}
\end{figure}
\end{document}

>>>>

We begin by benchmarking the performance improvement achieved by connecting
a previously published forwards-time simulation library,
\fwdpp{} \citep{fwdpp},
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

We implemented our method in \cpp{}, using \fwdpp{} library functions and
interface code (details below), using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{}, a Python package based
on \fwdpp.
% $\mu$ and $r$ were the means of Poisson processes, and mutations used the infinite-sites model \citep{Kimura1969-uc}.
% We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$.
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

@@

The total run times for simulations with selection are shown in Figure~\ref{fig:runtimes_selection}.  When tracking
neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for
these simulations).  Simulations with $N=5 \times 10^4$ timed out for region sizes larger than $10^3$.
Pedigree tracking
dramatically improves run times (right column of Figure~\ref{fig:runtimes_selection}).  For parameters that ran to completion when tracking neutral mutations and when tracking
the pedigree, the relative speedup due to pedigree tracking is up to $\approx 50$ fold
(Figure~\ref{fig:relative_speedup_selection}).  Pedigree tracking reduces runtimes more for larger
population sizes and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Note the log scale of the x axis in Figure~\ref{fig:runtimes_selection}, which
partially obscures an important observation that the run times appear approximately linear in region size when tracking
pedigree.  The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative
behavior with a numerically larger benefit to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

In our test runs, our method used substantially more RAM than simple forward simulations (see Appendix~\ref{ss:memuse}).
This is unsurprising:
unsimplified tree sequences grow quickly, and so storing history can use arbitrarily much memory.
However, this is not a requirement of the method, only a straightforwards consequence of a speed--memory tradeoff:
the amount of required memory is mostly determined by the interval between simplification steps,
but less frequent simplification reduces overall computation time (see Appendix~\ref{ss:gcinterval}).
In fact, our method could in some situations \emph{reduce} the amount of memory required,
if memory usage in the forwards simulation was dominated by the cost of maintaining neutral genetic variants.

\begin{figure}
    \includegraphics[]{sims/rawspeed}
    \caption{\label{fig:runtimes_selection}Total run time for a single simulation replicate as a function of region
        length, measured as the scaled recombination parameter $\rho = 4Nr$. The different colored lines represent
        different diploid population sizes ($N$). The left column shows run times for
        simulations including the tracking of neutral mutations. The right column of
panels shows run times tracking the pedigree instead of the neutral mutations. The dashed line in the upper right panel shows
the results for an implementation using \fwdpy{} where the pedigree simplification steps were handled in a separate thread of
execution and fitness calculations were parallelized.}
\end{figure}

@@

% Maybe an estimate of how long \emph{just} the pedigree recording and simplification takes,
% so that then we can say how fast the simulator would have to be to do $10^6$
% whole chromosomes for $10^7$ generations in a day.

@@

%%%%%%
\subsection*{Overview of \msprime\ APIs that enable the method}

@@

The \msprime\ Python Tables API provides a general purpose toolkit for importing
and processing succinct tree sequences. Interoperation with Python simulators
is then straightforward.  The implementation we benchmark here uses
\texttt{pybind11} (\url{https://github.com/pybind/pybind11/}) to inferface
with the \fwdpp{} \cpp{} API \citep{fwdpp}. No modifications were
required to the \fwdpp{} code base; rather, we simply need to bookkeep parent/offspring labels,
and perform simple processing of the recombination breakpoints from each mating
event to generate node and edge data. This information is then periodically copied
to the \msprime\ Tables API, where it is sorted and simplified.

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime\, we implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python (see Appendix~\ref{ss:simupop}).
For this proof-of-concept implementation, the simulation engine invokes
callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
It uses the \simupop{} library, interfacing with Python code (via our
package \ftprime{}) to simulate a discrete but large number of sites instead of the
infinite sites model; we provide results from this implementation alongside the
results from the \cpp implementation for simulations without selection  (see
Appendix~\ref{ss:timing_nosel}).
% $\mu$ and $r$ define per-base-pair mutation rate a
% the simulated sequence, in base pairs, as $L = \tfrac{\rho}{4N \mu}$.  We
% varied $N \in (1000, 10000)$, $\rho \in (10^3, 10^4)$.

\paragraph{Next steps.}
While these methods work well, and indeed result in significant performance
gains over non-ancestry based simulation, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in C or C++ out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime\ C library directly as part of
the core simulation logic, in the manner outlined in Algorithm W. If designed
in this way, the simulation engine would not need to keep copies of the node
and edge information and would not need to infer edges;
even greater gains in computational efficiency could then be obtained. To
facilitate this usage, a standalone, embeddable C library encompassing the core
tree sequence functionality is planned, which should make recording and
outputting tree sequences from any C/C++ code straightforward.

@@

@@

@@

The peak RAM use in gigabytes (GB) is shown for the simulations with selection in Figure~\ref{sfig:ramsel}.  For our parameters, which involved simplifying every
$10^3$ generations,  pedigree tracking requires more RAM.  This is because with pedigree tracking, a non-recombinant
individual still requires one new edge and two new nodes to be allocated.  Without pedigree tracking, no new memory is
allocated for such an offspring.  Thus, pedigree tracking accumulates a large amount of data in RAM until
simplification.  This extra RAM consumption is the trade off for reduced run times and the degree of extra RAM needed
can be reduced by simplifying more often.  The simulations where we run the simplification step in a separate thread
require even more RAM (dashed line in Figure \ref{sfig:ramsel}) because up to four sets of unsimplified data are
stored in RAM.  In practice, the RAM consumption can be minimized by simplifying more often.

@@

@@

We performed a limited set of simulations  to explore the effect of the simplification interval on run times and memory
use. Figure~\ref{sfig:gctime} shows that run times increase for simplification intervals less than 100 generations.  For
intervals $\geq 100$ generations, run times are very similar, with no appreciable difference between $10^3$ and $10^4$
generations.  As expected, the simplification interval has a near-linear effect on peak RAM use
(Figure~\ref{sfig:gcmem}).  Taken together Figures~\ref{sfig:gctime} and~\ref{sfig:gcmem} suggest that one can tune the
simplification interval to available RAM with little effect on run times provided that simplification doesn't occur too
often.

@@

We also performed a more limited set of simulations without natural selection.
For these we used \cpp{} as in the main text, but also provide results from our
proof-of-concept implementation that uses \simupop{} (see
Appendix~\ref{ss:simupop}).

The total run times are shown in
Figure~\ref{sfig:rawspeed_nosel} and for \cpp{} show the same qualitative behavior as simulations with selection
(Figure~\ref{fig:runtimes_selection}).  The relative improvement due to pedigree tracking is again substantial in these
simulations (Figure~\ref{sfig:speedup_nosel}).  In fact, we see more of a benefit to pedigree tracking here with \fwdpy{}
than we did with selection (Figure~\ref{fig:relative_speedup_selection}).  The reason is that the fitness function exits
close to instantly in simulations without selection that are based on \fwdpp{}, meaning that \fwdpy{} is doing little
more than generating random numbers and book-keeping in Figures~\ref{sfig:rawspeed_nosel}
and~\ref{sfig:speedup_nosel}.
For \simupop{}, the results show a qualitatively similar pattern but 1) the
simulations for $N = 10000$ or greater did no finish in the 72 hour limit or
exceed memory limits (more than 500 GB), and 2) the magnitude of the speedup is
much less.

The total RAM use for the
simulations without selection in Figure~\ref{sfig:ramnosel}.
Pedigree tracking used more RAM with \fwdpp{} and less
RAM with \simupop{}. The reason for reduced RAM use with \simupop{} is unclear;
with \fwdpy{}/\fwdpp{}, however the reason for the extra RAM use is as in
Appendix~\ref{ss:memuse}.

@@

\begin{figure}
    \includegraphics[]{sims/speedup_nosel}
    \caption{\label{sfig:speedup_nosel}Relative performance improvement due to pedigree tracking for simulations without
    selection.  Data points are taken from Figure~\ref{sfig:rawspeed_nosel} and show the ratio of run times tracking
neutral variation over the run times tracking the pedigree.}
\end{figure}
\begin{figure}
    \includegraphics[]{sims/memuse_nosel}
    \caption{\label{sfig:ramnosel}Peak RAM use in gigabytes (GB) for simulations without selection.  The plots are from
    the same data as in Figure \ref{sfig:rawspeed_nosel}.}
\end{figure}

\section{The \simupop{} implementation}
\label{ss:simupop}

We implemented simulations in a Python script, usable as a command-line utility,
taking input parameters for population size $N$ and scaled recombination rate $\rho$.
As described in 'Simulation benchmarks', we set scaled mutation rate $\theta=\rho$,
we used simplify interval of 1000 and simulated for $T=10N$ generations in each
simulation, introducing deleterious mutations at a rate of $\rho/100$ per
generation.
Because \simupop{} requires explicit locations for loci, we set the number of
loci $n=\rho / (4Nr)$ and for these simulations without selection marked $0$ of
these as selected loci.

\end{document}
